Data Specialization for a Staged Lambda Calculus

### 2015-02-27 paper/overview.tex
pass on overviet

\input{figures/quickselect-split}

It is well-known that for any given input list with $n$ elements, the
quick-select algorithm requires $\Theta(n)$ expected time (for any
second argument) if we select keys for the partition function
uniformly randomly or identically if we permute the input. Under this
assumption, our example program requires expected $\Theta(nm)$ time.

Suppose now that $m$ is large (close to $n$ for example).  In this
case, our program is essentially a quadratic-time algorithm, which can
be very slow even for modest values of $m$ and $n$. In this case, to
improve the run-time it appears that we have no choice but rewrite our
program to use a different algorithm.  One way to do this would be to
pre-sort the input list $l$ and then lookup the element with the
desired rank (as specified by the second argument).  While this
approach might sound sensible, attaining the desired performance
requires answering one more important question: what should the data
structure for storing the sorted input be?  Unfortunately using a list
does not work, because each look up in a list requires linear time on
average. One option is to use arrays but in this case, we would need
to implement a binary search to perform the lookup.  Another option is
to implement a balanced binary search tree data structure for storing
the elements in the list and then use balanced binary-search-tree
lookups, which are logarithmic time.  Using either one of these
techniques, we can reduce the run-time to $\Theta(n\log{n} + m\log{n})$,
which is significantly better that $\Theta(n^2)$.

In summary, to achieve the desired improvement, we wrote a two-pass
program, where the first pass sorts the input and the second pass
performs a binary search either on an array or a balanced tree. As is
probably apparent to the reader, the resulting program is both
algorithmically and structurally more complex: it requires conversions
between different data structures (list to array or binary search
tree) and requires implementing additional algorithms such as sorting
and binary search or possibly binary search trees.

%% As suggested by this relatively small example and experienced for
%% example by the graphics community, the design, implementation, and
%% maintenance of such multi-pass programs can be difficult due to the
%% increased complexity.

Instead of relying an writing multipass programs manually, in this
paper, we present techniques for automatically splitting staged
programs into multi-pass programs.  The key innovation behind our
approach is a {\em splitting algorithm} that splits a multistage
program into separate passes where each pass is a conventional program
expressed in a conventional functional language.  Our splitting
algorithm makes no restrictions on the input program: essentially any
functional program, including higher order programs, can be split.
Furthermore, the splitting algorithm is able to perform highly
non-trivial transformations on the program that can improve efficiency
significantly.

When applied to the quick-select example as described, our splitting
algorithm yields a two-pass program that uses the binary-search-tree
based implementation outlined above.  Specifically, in the first pass,
the program takes the input list and constructs a probabilistically
balanced binary search tree, which is isomorphic to a treap data
structure~\cite{treaps}.  In the second pass, the program performs,
for each rank, a binary search tree lookup, by walking the tree to
find the element with the desired rank.  The resulting program, shown
in \ref{fig:????}  takes the $O(n\log{n} + m\log{n})$.

To create the multi-pass algorithm, the splitting algorithm operates
by composing local transformations on the subterms of the input
program.  In particular, the algorithm has no special knowledge of the
quick-select algorithm, binary search trees, or how to perform lookups
on binary search trees, but it is able to derive all of these from the
input program.  But how?
%
The details of the algorithm are relatively involved but in the hope
that it aids understanding we present a brief informal overview of how
the algorithm operates.  
%
%% This exposition is necessarily imprecise but the algorithm is made
%% precise in \ref{sec:splitting} and implemented
%% (\ref{sec:implementation}{sec:examples}).
%

At a high level, the splitting algorithm operates on our example by
using the program code and the knowledge that the list (the first
argument) is first-stage and the rank ( the second argument) is second
stage, the algorithm then derives two functions: \texttt{qsStage1} and
\texttt{qsStage2} as follows.

\paragraph{Pass-1 code.}
The splitting algorithm scans the program and imitates an execution of
\texttt{qs} with just the first-stage input by evaluating as much code
as possible with the available first-stage values.  Specifically the
algorithm generates a function \texttt{qsStage1} to perform all the
recursive calls of \texttt{qs} and to evaluate all instances of the
partition function, which depend only on the input list.  The function
\texttt{qsStage1} produces a trace of the execution of \texttt{qs} on
its first input by generating a tree that collects the results from
all recursive calls along with a tag that indicates the control
branches taken.  The trace is represented as a tree because
\texttt{qs} has a binary control structure (casing on the list), which
leads to a binary tree data structure.  To ensure that the trace
contains the necessary results to complete the execution in the second
pass, where second-stage values may be used, the function
\texttt{qsStage1} includes in the nodes of the tree information such
as the key used for splitting the list at that recursive call.  The
resulting tree is thus a binary search tree, keyed by the ``pivot''
used by the partition.  In general we refer to the result data
structure of \texttt{qsStace1} as the {\em boundary}, as it passes
information from the first pass to the second.

\paragraph{Pass-2 code}
As the splitting algorithm scans \texttt{qs} for computations that can
be performed in the first stage and collects them into
\texttt{qsStage1}, it also collects computations that must be left to
the second stage in a separate function \texttt{qsStage2}.  This
function, which is executed in the second pass, takes as argument the
boundary and the second-stage argument (the rank) and performs a
lookup in the boundary data structure.  But how does \texttt{qsStage2}
knows to traverse the tree? It does not. Since the splitting algorithm
recorded the control flow of the first stage execution in the boundary
data structure, \texttt{qsStage2} simply follows this control
structure and performs at each point the parts of the computation from
\texttt{qs} that can now be performed in the presence of the
second-stage argument (the rank).  In the context of our example, this
performs a lookup on the boundary data structure by using the supplied
rank.

\paragraph{Summary.}
In summary, the splitting algorithm scans the program code for
first-stage computations (which depend only on first-stage values) and
separates them into function of the first pass. This function performs
the stage-1 computations and places the results into a boundary data
structure that both records the control flow and the results from the
first stage computation at each control-flow point.  As the splitting
function generates the code for the first pass, it also generates the
code for the second pass as a function that walks both the boundary
data (which essentially imitates execution) and at each control point
performs the part of the computation that depends on second-stage
values. We evaluate the resulting program by first evaluating the
function for the first pass and passing its return value (the
boundary) to the second pass, which then returns the desired result.

>>>>

\input{figures/quickselect-split-aligned}

The staged quickselect code shown in \ref{fig:qs-staged} makes
explicit the staging of all terms, making it natural to ask, whether
we can transform this code into manually implemented multi-pass code
described above and match the efficiency improvements of the manual
implementation. Considering the algorithmic knowledge and the
considerations needed, this may seem like a tall order.
Interestingly, our splitting algorithm achieves exactly this.

In the rest of this section, we present a brief, high-level overview
of the main ideas behind this algorithm using the quickselect example.
%
When applied to the staged code in \ref{fig:qs-staged}, our splitting
algorithm yields a two-pass program that uses the binary-search-tree
based implementation.  Specifically, in its first pass, the two-pass
program takes the input list and constructs a probabilistically
balanced binary search tree, which is isomorphic to a treap data
structure~\cite{treaps}.  In the second pass, the program performs,
for each rank, a binary search tree lookup, by walking the tree to
find the element with the desired rank.  
%
\ref{fig:qs-staged} illustrates the code for the first and second
passes \texttt{qs1} and \texttt{qs2} of the staged function
\texttt{qss}.

To create the multi-pass algorithm, the splitting algorithm operates
by composing local transformations on the subterms of the input
program.  In particular, the algorithm has no special knowledge of the
quickselect algorithm, binary search trees, or how to perform lookups
on binary search trees, but it is able to derive all of these from the
input program.
%

The splitting algorithm scans the program code for first-stage
computations (which depend only on first-stage values) and separates
them into the function, \texttt{qs1} of the first pass. This function
performs the stage-1 computations and places the results into a
boundary data structure that both records the control flow and the
results from the first stage computation at each control-flow point.
The function \texttt{qs1} performs all the recursive calls and
evaluate all instances of the partition function, which depend only on
the input list.  The function produces a boundary data structure that
collects the results from all recursive calls along with a tag that
indicates the control branches taken.  Since \texttt{qss} has a binary
control structure (casing on the list), the boundary is a binary tree.
To ensure that the trace contains the necessary results to complete
the execution in the second pass, where second-stage values may be
used, the splitting algorithm includes in the nodes of the tree
information such as the "pivot" used for splitting and the size of the
subtree at that note.  The resulting tree is thus not just a tree but
a binary search tree, keyed by the pivot and augmented with "size"
information.

As the splitting algorithm scans \texttt{qss} for computations that
can be performed in the first stage, it also collects computations
that must be left to the second stage in a separate function
\texttt{qs2}.  This function, which is executed in the second pass,
takes as argument the boundary and the second-stage argument (the
rank) and performs a lookup in the boundary data structure.  But how
does \texttt{qsStage2} knows to traverse the tree? It does not. Since
the splitting algorithm recorded the control flow of the first stage
execution in the boundary data structure, \texttt{qs2} simply follows
this control structure and performs at each point the parts of the
computation from \texttt{qss} that can now be performed in the
presence of the second-stage argument (the rank).  In the context of
our example, this performs a lookup on the boundary data structure by
using the supplied rank.

\begin{theorem}
  Consider an execution of \texttt{qs1} with a randomly permuted input
  of $n$ keys and performing $m$ executions of \texttt{qs2} with the
  result of \texttt{qs1}.  The total run-time for this computation is
  $\Theta(n\log{n} + m\log{n})$ in expectation.
\end{theorem}
